---
title: "Applying controls to samples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{applyingControls}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(flowFun)
```

If you have control samples, they may be used to calculate delta MFIs, which can then be used to plotting and differential expression testing. Doing so requires a few more steps than the basic ones outlined in the main workflow article. 

Before doing the steps we outline below, users should have a `GatingSet` corresponding to the primary data which the controls will be applied to. This `GatingSet` should be complete with gates created during pre-processing and clustering via the methods outlined in the main workflow article. Recall that `GatingSets` contain information including transformations, compensations, and gating schemes; this feature will be used to quickly process the data of the control files. 

### Reading and preprocessing control files

Like in the main workflow, the control files are read in with the function `load_cytoset_from_fcs()`. Instead of creating a fresh `GatingSet` with the data, however, we use the function `gh_apply_to_cs()`. This function takes the compensation matrices, transformations, and gates decided when pre-processing the main data and applies it to the control files, removing the need for the user to remember and reapply these steps themselves.

```{r preprocess-controls, eval=FALSE}
# Specify path to control FCS files
ctrl_dir <- "C:/path/to/controls"
# Read in files as cytoset
ctrl <- flowWorkspace::load_cytoset_from_fcs(list.files(ctrl_dir, full.names = TRUE), which.lines = 20000)

# Load saved GatingSet corresponding to main data
gs1 <- flowWorkspace::load_gs("path/to/gatingset")
# Apply transformations, compensations and gates to control gs
ctrl_gs <- flowWorkspace::gh_apply_to_cs(gs1[[1]], ctrl, compensation_source = "template") # make sure to exclude boolean
```

The result is a pre-processed `GatingSet`, similar to the original one, but now applied to the control files.

### Cluster controls

Before we can calculate delta MFIs, it is necessary to categorize the cells of the control files into the cell types identified in the main data. At this point there should be no need for the over-clustering and merging done in the main workflow; we simply use the final clustering from the main data and apply it to different data.

It is assumed that the main `GatingSet`, in this case `gs1`, contains a reference to an RDS file with information about the clustering. This should have been added with a call to `addClusteringToGatingSet()`; you can check that this is the case with `pData(gs1)`. 

Below we use the function `clusterControls()` to cluster the control files, which returns a `FlowSOM` object. We convert this to a `data.table` and save results with `addClustersToGatingSet()`.

```{r cluster-controls, eval=FALSE}
# Apply clustering to controls
# This may also be done manually using `FlowSOM::NewData`
fsom_projected <- clusterControls(ctrl_gs, gs1, "live")

# Get data.table for controls, and add clusters to corresponding GatingSet
ctrl_dt <- flowSOMToTable(fsom_projected)
addClustersToGatingSet(ctrl_dt, ctrl_gs, "live")
```

### Adding metadata to samples

In order to apply the controls to the data, the pipeline needs information about which sample each control file corresponds to. We incorporate this information with the function `addMetadataToGatingSet()`. To use it, you must have a table with sample information (see the vignette on preparing data for differential analysis for more details on how to create this table). 

The table must have at least a column `filename`, containing filenames of each sample, and a column(s) specifying control files and which sample file they correspond to. An example is shown below, where the `"FMO"` column lists the control files. 

:::{.border style="padding: 10px; border: 1px solid #dee2e6 !important;"}
```{r c1, echo=FALSE}
df <- data.frame(sample.name = c("Sample 1", "Sample 2", "Sample 3", "Sample 4", "Sample 5", "Sample 6"),
                 filename = c("sample_1.fcs", "sample_2.fcs", "sample_3.fcs", "sample_4.fcs", "sample_5.fcs", "sample_6.fcs"),
                 treatment = c("Treatment 1", "Treatment 2", "Treatment 2", "Treatment 1", "Treatment 2", "Treatment 1"),
                 FMO = c("T1_FMO.fcs", "T2_FMO.fcs", "T2_FMO.fcs", "T1_FMO.fcs", "T2_FMO.fcs", "T1_FMO.fcs"))

knitr::kable(df, 
             format = "html", 
             align = "l")
```
:::

Below we call `addMetaDataToGatingSet()`, which uses this table to add metadata to the `GatingSet`'s `pData()`. 

Note: adding this information to `pData(gs1)` allows the use of functions in the `ggcyto` package to facet by any group defined in the metadata.

```{r sample-data, eval=FALSE}
# Read in table of sample info
# Should have column named 'filename'
sample_info <- read.csv("path/to/sample/info/csv")
# Add metadata to GatingSet
addMetadataToGatingSet(gs1, sample_info)
# Check results (not run)
pData(gs1)
```

### Calculating delta MFIs

Finally, we can calculate delta MFIs with the function `gs_makeDeltaMFIs()`. It is necessary to specify which column of the metadata table (`pData(gs1)`) contains the names of the control files to use for this calculation. In this case it is done by setting `metadata_col = "FMO"`.

```{r mfis-and-testing, eval=FALSE}
# Get delta MFIs
delta_mfis <- gs_makeDeltaMFIs(gs1,
                               ctrl_gs,
                               subpopulations = subpops,
                               metadata_col = "FMO") # the name of the column containing control filenames
                                                     # (must be present in pData(gs1))
```


```{r extra-gates, eval=FALSE, include=FALSE}
#### Optionally, apply 1D boundary gates after clustering
# Plot channel marker densities by sample/metacluster
plot1DMarkerDensities(gs1,
                      channel = "FITC-A",
                      population = "live",
                      facet_by = "subpopulations", # may also facet by "samples"
                      inverse = FALSE)

# Add gate to GatingSet with openCyto
# Add boundary gate on channel; note arguments here are the same as column names for
#   the earlier gatingTemplate
openCyto::gs_add_gating_method(gs1,
                               alias = "+FITC-A",
                               parent = "9",         # add +FITC-A gate to metacluster 9
                               dims = "FITC-A",
                               gating_method = "gate_mindensity",
                               collapseDataForGating = TRUE,
                               groupBy = length(gs1))

```
