---
title: "Setting up differential analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{differentialAnalysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(flowFun)
```

In this article we focus only on the steps for preparing differential analysis, including how to specify sample information and define the comparisons you'd like to make.

### Specifying table of sample information

Once an appropriate `GatingSet` has been prepared, performing a basic differential analysis requires very little input from the user; however, it is crucial that this input is correct. In order to make comparisons between groups, R needs information about each sample in your experiment, and what group(s) they belong to. This information can then be used to generate a contrasts matrix and perform the tests. 

To facilitate this, the user must create a table that meets a few requirements. The first is that it contains at least 3 columns:

- `sample.name`, specifying the name of each sample.
- `filename`, specifying the file name of each sample. These names should correspond with the names you see when you call `sampleNames()` on your `GatingSet` or `flowSet`.
- A column specifying the experimental group to test within for differences. For example, `sex`, or `treatment`.

The second is that each sample contained in your `GatingSet` must be represented in the table. Meaning, if you have 15 samples in your `GatingSet`, you should have 15 corresponding rows in your table.

This table may be created programmatically in R, or exported from Excel as a CSV file. The figure below is an example of an appropriate table for an experiment with 4 samples.

#### Example 1

:::{.border style="padding: 10px; border: 1px solid #dee2e6 !important;"}
```{r ex1, echo=FALSE}
df <- data.frame(sample.name = c("Sample 1", "Sample 2", "Sample 3", "Sample 4"),
                 filename = c("sample_1.fcs", "sample_2.fcs", "sample_3.fcs", "sample_4.fcs"),
                 treatment = c("Treatment 1", "Treatment 2", "Treatment 1", "Treatment 2"))

knitr::kable(df, 
             format = "html", 
             align = "l")
```
:::

This basic example would allow for testing between `"Treatment 1"` and `"Treatment 2"`. Most experiments will have more information about their samples, and call for more particular comparisons. As long as the requirements listed above are adhered to, the user can add as many columns to this table as they like. The example below shows another example of an appropriate table, this time with more details.

#### Example 2 

:::{.border style="padding: 10px; border: 1px solid #dee2e6 !important;"}
```{r ex2, echo=FALSE}
df <- data.frame(sample.name = c("Sample 1", "Sample 2", "Sample 3", "Sample 4", "Sample 5", "Sample 6"),
                 filename = c("sample_1.fcs", "sample_2.fcs", "sample_3.fcs", "sample_4.fcs", "sample_5.fcs", "sample_6.fcs"),
                 treatment = c("Treatment", "Treatment", "Ctrl", "Treatment", "Ctrl", "Treatment"),
                 age.group = c("<65", "65+", "65+", "65+", "<65", "<65"),
                 sex = c("female", "female", "male", "male", "female", "male"),
                 pregnancy = c("pregnant", "not.pregnant", "X", "X", "not.pregnant", "X")
                 )

knitr::kable(df, 
             format = "html", 
             align = "l")
```

:::

This table corresponds to an experiment with 6 samples, and has 3 more columns than the previous example: `age.group`, `sex`, and `pregnancy`. Note that in the column `pregnancy`, possible values are `"pregnant"`, `"not.pregnant"`, or, when `sex == "male"`, `"X"`. Use an `"X"` (or leave the cell empty, if making the table in Excel) when a category is not applicable to a sample. 

Many more comparisons can be created with this table, e.g. males vs. females within the control group only, or pregnant vs. non-pregnant females within the treatment group (of course these comparisons wouldn't be especially helpful with so few samples, but this table is for demonstration purposes only).

### Define comparisons object

The other essential input expected from the user is a list defining the comparisons to make between groups. A summary of how to define this object can be found in the examples for the function `prepareSampleInfo()` (which may be brought up by typing `?prepareSampleInfo` into the console when `flowFun` has been loaded with `library(flowFun)`). Here we will give a few more thorough examples.

The comparisons must be a named list of named lists. This description isn't too helpful on its own, so we will first give a basic example, then elaborate. The chunk below defines two basic comparisons using the previously shown table (Example 2).

```{r comparisons1}
comparisons1 <- list(
  old_vs_young = list(age.group = list("<65", "65+")), # comparison 1
  male_vs_female_tmnt = list(treatment = "Treatment", sex = list("female", "male")) # comparison 2
)
```

When defining a named list in R, the left-hand side (LHS) of each `=` is the name of that item. So, in this example, `comparisons1` has two elements named `"old_vs_young"` and `"male_vs_female_tmnt"`, each one corresponding to a test we'd like to make. 

Note that these names may be anything you like, but special characters like `#` and `?` should be avoided. It is also recommended that underscores `_` or periods `.` are used instead of spaces, to avoid unexpected behavior from certain functions.

Then for each comparison, we have another named list. This is where we instruct which groups should be included in that test. In this case, these lists are `list(age.group = list("<65", "65+"))`, and `list(treatment = "Treatment", sex = list("female", "male"))`. Here we must be more careful about list names. The names must also be column names in your table of sample info, which in this case are `age.group`, `treatment`, and `sex`. This tells R which columns to check to find the info it needs to make that particular comparison. 

Finally, for each column/variable relevant to that comparison, we specify which group(s) are of interest. For comparison 1, by setting `age.group = list("<65", "65+")` we indicate that we'd like to test for differences between old and young groups (referencing the table above, this would be sample 1, 5, and 6 vs. 2, 3, and 4). 
:::{.border style="padding: 10px; border: 1px solid #dee2e6 !important;"}
```{r c1, echo=FALSE}
df <- data.frame(sample.name = c("Sample 1", "Sample 2", "Sample 3", "Sample 4", "Sample 5", "Sample 6"),
                 filename = c("sample_1.fcs", "sample_2.fcs", "sample_3.fcs", "sample_4.fcs", "sample_5.fcs", "sample_6.fcs"),
                 age.group = c("<65", "65+", "65+", "65+", "<65", "<65")
                 )

knitr::kable(df, 
             format = "html", 
             align = "l")
```

:::

For comparison 2, by setting `treatment = "Treatment"`, we indicate that we only want to test within the treatment group, and by setting `sex = list("female", "male")` we indicate that within that group we want to test for differences between males and females (this would be sample 1 and 2 vs. 4 and 6). 

:::{.border style="padding: 10px; border: 1px solid #dee2e6 !important;"}
```{r c2, echo=FALSE}
df <- data.frame(sample.name = c("Sample 1", "Sample 2", "Sample 4", "Sample 6"),
                 filename = c("sample_1.fcs", "sample_2.fcs", "sample_4.fcs", "sample_6.fcs"),
                 treatment = c("Treatment", "Treatment", "Treatment", "Treatment"),
                 sex = c("female", "female", "male", "male")
                 )

knitr::kable(df, 
             format = "html", 
             align = "l")
```

:::
