---
title: "flowFun"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flowFun}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Installation
```{r, eval = FALSE}
install.packages("devtools", dependencies = TRUE, build_vignettes = TRUE)
devtools::install_github("00berst33/flowFun")
```

# Reading in Data

Cytometry data often contains millions of cells, so to facilitate fast and efficient manipulation of these large datasets, this package encourages the use of the `data.table` data structure.

Other advantages of using this data structure is that it is futureproof, as it can easily be converted to a `data.frame`, a structure found in base R. This also makes the package more flexible, as although it focuses on the use of the FlowSOM package for clustering, the package could easily be expanded to work with any other clustering algorithm the user may prefer.

`getTableFromFCS()` allows the user to read .fcs files directly into a data table, either by specifying a list of directories containing relevant files, or by providing a list with the relative or absolute filepaths of each file of interest. The expression data for each sample are joined by row, and two columns are joined to the left of the table: `.id` and `cell_id`.

By default, all cells are sampled from each .fcs file of interest, but the parameter `num_cells` may be specified to select 

```{r load-data}
library(flowCore)
library(tidytable)
library(flowFun)

# Specify path to .fcs file
file <- system.file("extdata", "aggregate.fcs", package = "flowFun")
# Read in as tidytable
table <- getTableFromFCS(file, num_cells = 1000)
```

The following is a snippet of the resulting table:

```{r print-table, fig.align='center', echo=FALSE} 
table$.id <- basename(table$.id)

table %>% 
  tidytable::select(tidytable::all_of(1:10)) %>%
  tidytable::slice_head(5) %>%
  data.frame() %>%
  knitr::kable(
    format = "html",
    align = "c",
    linesep = "",
    )
```

# Preprocessing

Once setup files have been cleaned and a compensation matrix has been generated, the next step is to preprocess the raw data. In order, the steps this pipeline implements are as follows:

1. Removing margin events (i.e. values that fall outside the range the cytometer should be able to pick up).
2. Removing doublets.
3. Removing debris.
4. Compensating the data.
5. Transforming the data.
6. Removing dead cells.
7. Quality control via flowCut, to check and correct for occurrences like clogs, speed changes, etc. and flag any files that may be too low-quality to include in the analysis.
      
Each these steps, other than acquiring a compensation matrix, may be left entirely automated if desired. However, oftentimes parameter tuning is necessary for optimal results. Furthermore, some samples' data may be distributed in a way that is markedly different from the average, making automated gating ineffective. For this reason the user also has the option of manually drawing gates, by either simply using the same one for each sample, or retroactively selecting samples which the automated gating performed poorly for and re-drawing gates as necessary. Although it is implemented, it is highly recommended to avoid as much manual gating as possible. Doing so greatly improves the objectivity and reproducibility of the analysis.

Note that this workflow is currently restricted to the outlined gating scheme, but will later support user-defined gating schemes. ###

The easiest way to perform preprocessing is via the function `doPreprocessing()`. In the example below, each possible parameter is listed. As input, the function accepts either a directory name, an array specifying paths to files of interest, or a `data.table` resulting from `getTableFromFCS()`. The user may then provide their compensation matrix, if they have one. A transformation may be specified by simply choosing one of the enumerated options under `transformation_type`, but the user may also give a custom transformation if preferred. 

If your panel includes a viability stain, it is necessary for the channel corresponding to it to be specified by `ld_channel`, and a compensation matrix and appropriate transformation to be provided, so that dead cells may be gated out accurately. If any of these are not present, the function will not apply a live/dead gate.

The strictness of the doublet removal may be controlled with the `nmad` parameter. Decreasing it results in a more strict gate, and vice versa.

If `debris_gate` and `live_gate` are left unspecified, then automated gates will be found for each sample. Alternatively, the user may specify their own gate for each of these parameters, and the same gate will be applied across all samples. These gates should be defined via either the flowCore or flowDensity packages, as demonstrated below.

```{r manual-gates, fig.height=5, fig.width=7}
library(flowDensity)

# Get filepath to .fcs file of interest
file <- system.file("extdata", "sample_2.fcs", package = "flowFun")

# Read in compensation matrix
comp_mat <- system.file("extdata", "compensation_matrix.csv", package = "flowFun")
comp_mat <- read.csv(comp_mat, check.names = FALSE)

# Preview parameters
previewPreprocessing(file,
                     ld_channel = "BUV496-A",
                     compensation = comp_mat,
                     transformation_type = "logicle")
```

Here we see that the gates for dead cells and doublets are a bit loose for this sample. As an easy fix, we may adjust `nmad`, and draw a simple manual gate. A rectangle gate is used for this example, but ellipsoid and polygon gates via the flowCore package are also supported.

```{r manual-fix, fig.height=5, fig.width=7, fig.align='default'} 
# Define manual gate for removal of dead cells
manual_live <- flowCore::rectangleGate(.gate = list("BUV496-A"= c(-1, 1.7),
                                                    "FSC-A" = c(0, 270000)))

# Check new plots
previewPreprocessing(file,
                     ld_channel = "BUV496-A",
                     compensation = comp_mat,
                     transformation_type = "logicle",
                     nmad = 3,
                     live_gate = manual_live)
```

Once the desired parameters have been selected, all samples may be processed. During this step, the user can set minimum thresholds a sample must meet to avoid being removed from analysis. `pctg_live` sets the minimum proportion of live cells a sample must have, and `pctg_qc` sets the minium proportion of quality events a sample must have. To avoid any samples being removed, these parameters may be set to 0.

```{r preprocess-samples, eval=FALSE}
# Get all sample data
table <- system.file("extdata", "raw_samples.rds", package = "flowFun")
table <- readRDS(table)

# Preprocess all samples
prepr_table <- doPreprocessing(table,
                               num_cells = 10000,
                               ld_channel = "BUV496-A", 
                               compensation = comp_mat,
                               transformation_type = "logicle",
                               live_gate = manual_live,
                               nmad = 3.5, 
                               pctg_live = 0.7,
                               pctg_qc = 0.8,
                               save_plots = FALSE,
                               save_fcs = FALSE)
```

Once the processing is finished, the script will output files to check for any mistakes or concerns. The first is a pdf highlighting the cells that were removed in each .fcs file. The others are plots for any files that were flagged during quality control, where each
channel is plotted against time, and removed events are marked. If desired, the names of the resulting directories and files may be changed, and the preprocessed .fcs files can be written to disk.

# Clustering

This package is compatible with various clustering algorithms, but was designed specifically with FlowSOM in mind due to its fast runtime and tools for data visualization. Therefore, much focus will be given to FlowSOM in the following section.

The strategy for identifying cell type populations outlined by this workflow is not entirely automated, and instead uses a human-in-the-loop approach. Such an approach aims to reduce the burden of analyzing such a large volume of data, while still incorporating expert knowledge to ensure that results are biologically meaningful. To accomplish this, the data is initially overclustered. The expert then examines the result of the clustering through various relevant plots, taking into account not only marker expression but also cluster sizes and mathematical distance, and uses their best judgement to merge clusters until an appropriate final clustering has been reached and labelled. 

It is also necessary for the user to specify which markers are to be used for clustering. There are a few considerations to account for when selecting these, the most important being that it is highly recommended that any markers the user intends to test for differential expression are NOT used for clustering. There is a number of reasons for this, but in short, it generally makes little sense to test for differential expression of a marker in a cluster when the cluster is defined by high or low expression of said marker (i.e. testing for differential expression of CD8 in CD8 T cells). Furthermore, parameters like side and forward scatter should not be included, and if a viability stain was used, it should also be excluded.

Below, the preprocessed data from earlier is clustered using the FlowSOM algorithm via the function `flowSOMWrapper()`. This outputs the same table that is given to it, but with the columns `Metacluster` and `Cluster` joined to the right, giving the results of the clustering. Specifying a seed is optional, but strongly recommended for the sake of reproducibility.

``` {r flowsom-clustering}
# Read in table
prepr_table <- system.file("extdata", "prep_table.rds", package = "flowFun")
prepr_table <- readRDS(prepr_table)

# Define markers/columns to use for clustering
cols_to_cluster <- c(12, 14:16, 18, 20:25, 27:34, 36)

# Perform clustering
fsom_dt <- flowSOMWrapper(prepr_table,
                          cols_to_cluster = cols_to_cluster,
                          num_clus = 25,
                          xdim = 12,
                          ydim = 12,
                          seed = 42)

```

```{r print-fsom-table, echo=FALSE} 
fsom_dt %>% 
  tidytable::select(tidytable::all_of(c("cell_id", "Metacluster", "Cluster"))) %>%
  tidytable::slice_head(5) %>%
  data.frame() %>%
  knitr::kable(
    format = "html",
    align = "c",
    linesep = "",
    )
```

A number of different plots may be generated by this package to aid the user in merging metaclusters, but the arguably the most important are heatmaps and dimension reduction plots. Below, a heatmap of MFIs by metacluster is generated. 

```{r first-heatmap, fig.height=5, fig.width=6, fig.align='center'}
# Generate heatmap
plotMetaclusterMFIs(fsom_dt)
```

This heatmap is what the user should be using as their main reference when merging metaclusters. The dendrogram on the left-hand side, which displays a hierarchal clustering of the metaclusters, is of particular interest, and this hierarchy may be followed to perform the clustering. However, this dendrogram should not be followed blindly. First, the metaclusters that are determined to be similar by the clustering may not be of biological interest; some cell type markers may be of greater significance than others, and this is something that the clustering does not take into account. Second, dendrograms may be created with different linkage methods (single-linkage, average-linkage, centroid-linkage, etc.), and these different methods may result in slightly different hierarchies. This is not to say that the dendrogram is useless, only that it has its limitations.

The user should also consider cluster size when merging. If a metacluster is rather small, and differs from its neighboring clusters in irrelevant markers, then it is reasonable to merge it. However, if the metacluster is small but distinct in markers of interest, it may be left
unmerged. Finally, if a metacluster does not represent any particular cell type of interest, it may be dropped from the analysis entirely. For example, depending on how strict you were with your gating, there may be some debris remaining in your preprocessed data. The clustering algorithm will likely cluster these together, and you may decide to exclude them from further analysis.

To further aid the decision making process, a UMAP colored by metacluster is generated below. 

```{r umap, fig.height=5, fig.width=6, fig.align='center'}
# Generate UMAP
plotUMAP(fsom_dt, num_cells = 5000, seed = 42)

```

This plot is most useful to check metaclusters' relative sizes, and how similar they are to one another. The closer clusters (and cells) are to each other, the more similar they are to each other. However, the opposite is not necessarily true.

By examining these plots, it starts to become clear which metaclusters are more similar to one another, and what their phenotype may be. For example, we see that metaclusters 1, 2, and 9 are all closely linked by the dendrogram in the heatmap, and are rather close to one another in the UMAP. They all have a high expression of the markers CD3 and CD8, but differ noticeably in markers PD1, CCR7, and CD45RA. It seems that these metaclusters are each a subset of CD8 T cells, and it is reasonable to merge them. Depending on the question being investigated, the user may wish to instead keep them separate, and perhaps specify exhausted and non-exhausted CD8 T cells, which could also be reasonable. 

It should be noted that if a user does desire to look at a particular cell type in greater resolution, this workflow allows for backgating on and reclustering cells belonging to a particular metacluster; therefore identifying every sub-population of interest is not necessary at this step. For now, this example will find an initial low-resolution clustering.

The function `editTableMetaclusters()` can be used to merge and rename metaclusters. A new column, `Meta_original`, is added to the given table, specifying the metaclusters that were found when the clustering algorithm was first called. The existing `Metaclusters` column is edited to contain the new assignments. 

```{r first-merge}
# Merge metaclusters
fsom_dt <- editTableMetaclusters(fsom_dt, 
                                 new_labels = c("9" = "1",
                                                "2" = "1",
                                                "8" = "4",
                                                "23" = "15",
                                                "21" = "15",
                                                "20" = "15",
                                                "22" = "15",
                                                "7" = "6"))

```

```{r print-fsom-table-merge1, echo=FALSE} 
fsom_dt %>% 
  tidytable::select(tidytable::all_of(c("cell_id", "Meta_original", "Metacluster", "Cluster"))) %>%
  tidytable::slice_head(5) %>%
  data.frame() %>%
  knitr::kable(
    format = "html",
    align = "c",
    linesep = "",
    )
```

A new heatmap and UMAP are generated to reflect the new clustering, and help decide whether any further merging should be performed.

```{r second-plots, fig.height=5, fig.width=6, fig.align='center'}
# Generate new heatmap
plotMetaclusterMFIs(fsom_dt, cols_to_cluster)

# Generate new UMAP
plotUMAP(fsom_dt, num_cells = 5000, seed = 42)
```

Again, metaclusters are merged and new plots are generated.

```{r second-merge, fig.height=5, fig.width=6, fig.align='center'}
# Merge metaclusters
fsom_dt <- editTableMetaclusters(fsom_dt, new_labels = c("12" = "5",
                                                         "13" = "5",
                                                         "10" = "6",
                                                         "14" = "6",
                                                         "25" = "18",
                                                         "16" = "11",
                                                         "24" = "3"))

# New heatmap
plotMetaclusterMFIs(fsom_dt, cols_to_cluster)

# New UMAP
plotUMAP(fsom_dt, num_cells = 5000, seed = 42)
```

Finally, the remaining metaclusters may be labelled by setting each of them equal to the desired name, instead of another metacluster to be merged into.

```{r new-labels, fig.height=5, fig.width=6, fig.align='center'}
# Label metaclusters
fsom_dt <- editTableMetaclusters(fsom_dt, new_labels = c("6" = "Monocytes",
                                                         "5" = "cDC",
                                                         "4" = "B cells",
                                                         "17" = "NK T cells",
                                                         "19" = "NK cells",
                                                         "1" = "CD8 T cells",
                                                         "11" = "gdT cells",
                                                         "15" = "CD4 T cells",
                                                         "18" = "pDC",
                                                         "3" = "Undefined"))

# Generate final heatmap
plotMetaclusterMFIs(fsom_dt, cols_to_cluster)
```

An annotated heatmap displaying the merged clusters may be created with the function `annotateMFIHeatmap()`.

```{r annotate-heatmap, fig.height=5, fig.width=6, fig.align='center'}
# Generate annotated heatmap
annotateMFIHeatmap(fsom_dt, cols_to_cluster)
```

### Reclustering

This script also allows the user to perform backgating and reclustering on any identified cell types of particular interest. This clustering may be done using either a selected number of principal components obtained from PCA, or a new subset of markers defined by the user. The former approach is demonstrated below to further investigate the cells identified as CD8 T cells.

```{r subset}
# Get table of only CD8 T cells
new_table <- createFilteredAggregate(fsom_dt, 
                                     num_cells = Inf, 
                                     metaclusters = "CD8 T cells",
                                     clusters = NULL)
```

```{r subset-table, echo=FALSE} 
new_table %>% 
  tidytable::select(tidytable::all_of(c("cell_id", "Meta_original", "Metacluster", "Cluster"))) %>%
  tidytable::slice_head(5) %>%
  data.frame() %>%
  knitr::kable(
    format = "html",
    align = "c",
    linesep = "",
    )
```

One an appropriate subset has been created, the next step is to create a scree plot, which may be used to determine which principal components should be used in the analysis. Typically, there is a distinct "elbow" in the plot, where the amount of variance explained by each component becomes drastically smaller. So, if this elbow appears at principal component 5, components 1-5 would be selected to recluster the data. If there is no obvious elbow, it is instead reasonable to use whichever number of components explains roughly 70-80% of the variance. First, `doPCA()` is called to perform PCA. Then, `plotPCAScree()` creates the scree plot.

```{r pca, fig.height=4, fig.width=6, fig.align='center'}
# Perform PCA
pca_obj <- doPCA(new_table, cols_to_cluster)

# Draw scree plot
plotPCAScree(pca_obj)
```
Here, there appears to be a fairly distinct elbow at `M = 6`. To recluster using these principal components, the following call to `clusterSubsetWithPCA()` is made, with `num_components = 6`. Fewer metaclusters are created than in the initial clustering, as fewer cell types are expected.

```{r reclustering}
# Recluster CD8 T cells
fsom_sub <- clusterSubsetWithPCA(new_table, 
                                 pca_obj = pca_obj, 
                                 num_components = 6, # number of principal components
                                 num_clus = 15, # number of metaclusters
                                 seed = 33)
```

Note that the plot functions demonstrated earlier function the same for any subsets created. Since the current cells of interest are CD8 T cells, it is appropriate to define a new subset of markers to create the heatmap.

```{r reclustering-plots, fig.height=6, fig.width=5, fig.align='center'}
# Define new columns to use
cols_of_interest <- c(15, 16, 18, 23:25, 28, 32)

# Generate heatmap for subsetted cells
plotMetaclusterMFIs(fsom_sub, cols_of_interest)

# Likewise, generate UMAP
plotUMAP(fsom_sub, num_cells = 5000, seed = 33)
```

## Plots

Various plotting functions that are not essential to the workflow, but potentially still of interest, have been included in this package. 

- searchByExpression()
- plotLabeled2DScatter()
- plotSampleProportions()
- plotClusterGroupProportions()
- plotClusterFileProportions()


# Differential Analysis

The final analysis step in this workflow is differential analysis. Differential marker expression is implemented using limma, and differential count analysis using edgeR, both R packages found on Bioconductor. 

Some preparation is necessary before any tests can be performed. First, it is necessary to specify information about each sample, including any FMO or isotype controls, and which groups of interest it belongs to (e.g. control vs. treatment). This information may be provided as a .csv file or `data.frame` to the function `prepareSampleInfo()`, which will prepare the table for further use by the workflow and ensure that its values are R-friendly (no special characters like @ or #, no empty values, etc.). Below is an example of what an appropriate .csv file might look like.

```{r sample-info, echo=FALSE} 
# Get filepath
file <- system.file("extdata", "sample_info.csv", package = "flowFun")
table <- read.csv(file)

table %>%
  knitr::kable(
    format = "html",
    align = "c",
    linesep = "",
    )
```

Before calling `prepareSampleInfo()`, the user must also specify the comparisons they wish to make between groups, as a series of nested lists. Each nested list should include at least two levels of a factor to be used for the comparison. Each comparison may be named however the user desires, given that it is R-friendly. Below, two comparisons, `ctrl_vs_mibc` and `nac_vs_no_nac` are defined.

```{r comparisons}
comparisons <- list(
  ctrl_vs_mibc = list(disease = list("MIBC", "Ctrl")),
  nac_vs_no_nac = list(disease = "MIBC", NAC = list("NAC", "No.NAC"))
)
```

The factor names (which would be "disease" and "NAC" in the above example), unlike the names of comparisons, should be identical to their column names in the .csv file or data frame given to `prepareSampleInfo()`. Similarly, the factor levels (which would be "Ctrl", "MIBC", "NAC", and "No NAC" in the above example), should be listed identical to how they appear in the data frame. Note how these values match those present in the table just shown.

Once these two items have been specified, the remainder of analysis is rather straightforward. `prepareSampleInfo()` is called, formatting the given table as necessary, and adding a new column called `group`.

```{r prep-sample-info}
# Get filepath
file <- system.file("extdata", "sample_info.csv", package = "flowFun")

# Prepare metadata for further analysis
sample_info <- prepareSampleInfo(file, 
                                 name_col = "sample.name", # column containing sample names
                                 filename_col = "filename", # column containing filenames
                                 comparisons = comparisons)
```

```{r print-sample-info, echo=FALSE}
sample_info %>%
  knitr::kable(
    format = "html",
    align = "c",
    linesep = "",
    )
```

Next, the design, contrasts, and count matrices are created. The current population of interest (in this case, `fsom_dt`) should be given to `makeCountMatrix()`. Furthermore, the parameter `meta_names` should be used to specify the metaclusters to use for analysis. By default, all metaclusters are included, but if the final merging included an undefined metacluster, or some are of no interest, it is useful to specify this parameter.

```{r matrices}
# Generate design matrix
design <- makeDesignMatrix(sample_info)

# Generate contrasts matrix
contrasts <- makeContrastsMatrix(sample_info, comparisons)

# Define metaclusters to test - here, "Undefined" is excluded
meta_of_interest <- c("Monocytes", "CD4 T cells", "CD8 T cells", "gdT cells", 
                      "B cells", "pDC", "cDC", "NK cells", "NK T cells")

# Generate matrix of sample/metacluster cell counts
counts <- makeCountMatrix(fsom_dt, 
                          meta_names = meta_of_interest,
                          min_cells = 3, 
                          min_samples = 4)
```

The table of counts is printed below:

```{r print-counts, echo=FALSE}
counts %>%
  knitr::kable(
    format = "html",
    align = "c",
    linesep = "",
    )
```

#### Differential Abundance

Once these matrices have been acquired, to perform differential abundance analysis, the user only needs to pass them to `doDAAnalysis()`, and specify a normalization method, if any. Results are returned as a list, where each entry is a table corresponding to a comparison. The tables are named after the comparison they correspond to, and contain results of likelihood ratio tests performed by edgeR, with each test being ranked by its adjusted p-value. 

Results are also saved as a .csv file for each comparison in the directory `Analysis Results/edgeR`. If running an edgeR analysis on multiple objects, or running the script multiple times for any reason, the user may wish to create sub-directories within `Analysis Results/edgeR` to stay organized. To do this, simply set the \code{dir_tables} parameter to a preferred sub-directory name.

```{r da-analysis}
# Perform differential abundance analysis
da_results <- doDAAnalysis(design = design, 
                           counts = counts, 
                           contrasts = contrasts,
                           sample_df = sample_info, 
                           norm_method = "TMM")
```

The resulting individual tables may be accessed with either double brackets `[[]]` or `$`, e.g. `da_results[[1]]` or `da_results$MIBC_NAC__vs__MIBC_No.NAC`. In this example, two comparisons were tested, so the resulting list contains two tables:

```{r da-analysis-tables, echo=FALSE}
da_results[[1]] %>%
  knitr::kable(
    format = "html",
    align = "c",
    caption = "Ctrl vs. MIBC",
    linesep = "",
    )

da_results[[2]] %>%
  knitr::kable(
    format = "html",
    align = "c",
    caption = "NAC vs. No NAC",
    linesep = "",
    )
```

#### Differential Expression

To test for differential expression, the design, contrasts, and count matrices must again be specified, in addition to the markers to be tested. Recall that in most cases, these markers should not be the same as those used for clustering.

```{r de-analysis}
# Perform differential expression analysis
de_analysis <- doDEAnalysis(fsom_dt, 
                            sample_df = sample_info, 
                            design = design, 
                            contrasts = contrasts, 
                            counts = counts, 
                            markers_of_interest = c("PHA-L", "IL10R"),
                            meta_names = meta_of_interest)

# cols_from_fcs had to be retroactively changed
```

```{r extra, include=FALSE}
# 12 - cDC
# 10 - monocytes
# 14 - monocytes
# 13 - cDC
# 6 - monocytes
# 5 - cDC
# 4 - B cells 
# 18 - pDC
# 17 - NK T cells
# 19 - NK cells
# 11 - gdT cells
# 3 - undefined
# 1 - cd8 T cells
# 16 - gdT cells
# 15 - cd4 T cells
# 25 - pDC
# 24 - undefined
```
